Machine Learning Tools for Open Cluster Characterization with Gaia DR2 Data

### 2020-12-31 thesis/thesis.tex
thesis: Apply supervisor suggestions
Co-Authored-By: Cesar Guzmann <15159632+cguz@users.noreply.github.com>

%---------------------------
% Título del trabajo y autor
%---------------------------
\title{Using Machine Learning Tools for Open Cluster Characterization with Gaia DR2 Data}
\author{Carlos David Álvaro Yunta}
\date{12th of November, 2020}
\director{César Augusto Guzmán Álvarez}
\nombreciudad{Madrid, Spain}

@@

The characterization and knowledge of \emph{Open Clusters} allows us to better understand properties and mechanisms about the Universe
such as stellar formation and the regions where those events occur. They also provide information about stellar processes and the
evolution of the galactic disk.

In this work, we develope a method to characterize those clusters by using \emph{artificial intelligence} tools like clustering
by \emph{K-Means}, or clustering based on \emph{Artificial Neural Networks} by implementing the \emph{Deep Embedded Clustering}
model. We are using \emph{Gaia DR2 database} as data source for testing our models.

The developed method aims to improve the existing ones both in terms of \emph{computational efficiency}, with lower computational requirements,
and \emph{ease of use}, by reducing the number of hyperparameters to configure in order to obtain a good characterization of the analyzed clusters.

As detailed below, our method achieves good results, becoming even better in some cases when the results are compared with other current methods.

@@

{\bf Key words.} open clusters characterization --- machine learning --- gaia dr2 --- data analysis --- deep embedded clustering

@@

{\bf Palabras Clave:} caracterización de cúmulos abiertos --- inteligencia artificial --- gaia dr2 --- análisis de datos

@@

Stellar open clusters (OCs) \cite{janes1982open} are groups of stars gravitationally bound originated from a single molecular gas cloud.
Thus they share the same chemical composition and age, and they have similar relative positions and proper motion.
These astronomical objects are of fundamental importance to understand the spiral structure,
the dynamics and the chemical evolution of our galaxy.

Although most stars in the Milky Way are presented isolated, it is considered that most of them (or even all)
are formed in clustered environments and spend a period of time gravitationally bound with their siblings embedded
in their original molecular cloud \cite{clarke2000theformationofstellarclusters} \cite{portegies2010young}.
The evolution of these systems tends to sparse them in a few million years by interacting gravitationally with other systems.
Galactic tidal forces and mechanisms that involves gas loss driven by stellar feedback are other causes of disruption
\cite{brinkmann2017bound}.
Nevertheless, a small fraction of these systems will survive in the initial state and persist bound in bigger timescales.

@@

Stars that belong to a same OC share relative positions, inherited from their original gas cloud.
So their distance to the Earth is similar for all of them and, therefore, they have a narrow dispersion in their parallax value.
They also share similar values of proper motion, both in right ascension and declination.
Another property they share is their chemical composition.
Thus, their metallicity must be uniform, since these stars were born from the same gas cloud and at the same time stage.
However, to take this last property into account, we are faced with the drawback that this parameter is poorly reported in Gaia DR2 database.

@@

The main aim of this thesis is to \emph{build an unsupervised clustering model for open cluster characterization}.

We want this model to be \emph{non-supervised non-parameterized} in order to fit a wide range of clusters without
the need of fine tunning hyperparameters regarding the topology or size of the studied clusters.

@@

\section{Comparison}

>>>>

%---------------------------
% Título del trabajo y autor
%---------------------------
\title{Machine Learning Tools for Open Cluster Characterization with Gaia DR2 Data}
\author{Carlos David Álvaro Yunta}
\date{31st of December, 2020}
\director{Dr. César Augusto Guzmán Álvarez}
\nombreciudad{Madrid, Spain}

@@

The characterization and knowledge of \emph{Open Clusters} allow us to better understand properties and mechanisms about the Universe
such as stellar formation and the regions where those events occur.
They also provide information about stellar processes and the evolution of the galactic disk.

In this work, we develop a method to characterize open clusters by using \emph{artificial intelligence} tools,
such as clustering by \emph{K-Means} and clustering based on \emph{Artificial Neural Networks} by implementing the \emph{Deep Embedded Clustering} model.
We are using \emph{Gaia DR2 database} as data source for testing our models.

The developed method aims to improve current techniques of the state of the art.
We achieve improvements not only in terms of \emph{computational efficiency}, with lower computational requirements, but in \emph{usability},
reducing the number of hyperparameters to obtain a good characterization of the analyzed clusters.

Our method achieves good results, becoming even better in some cases when the results are compared with current techniques.

@@

{\bf Key words.} open clusters characterization, machine learning, gaia, data analysis, deep embedded clustering

@@

{\bf Palabras Clave:} caracterización de cúmulos abiertos, inteligencia artificial, gaia, análisis de datos

@@

Stellar open clusters (OCs) \cite{janes1982open} are groups of stars gravitationally bound originated from a single molecular gas cloud.
They share the same chemical composition and age. Moreover, they have similar relative positions and proper motion.
Those astronomical objects are relevant to understand the spiral structure, the dynamics and the chemical evolution of our galaxy.

Although most stars in the Milky Way are presented isolated, it is considered that most of them (or even all)
are formed in clustered environments and spend a period of time gravitationally bound with their siblings embedded
in their original molecular cloud \cite{clarke2000theformationofstellarclusters} \cite{portegies2010young}.
The evolution of those systems tends to sparse them in a few million years by interacting gravitationally with other systems.
Galactic tidal forces and mechanisms that involves gas loss driven by stellar feedback are other causes of disruption
\cite{brinkmann2017bound}.
Nevertheless, a small fraction of those systems will survive in the initial state and persist bound in bigger timescales.

@@

Stars that belong to the same OC share relative positions, inherited from their original gas cloud.
Which means that their distances to the Earth are similar for all of them and, therefore, they have a narrow dispersion in their parallax value.
They also share similar values of proper motion, both in right ascension and declination.
Another property they share is their chemical composition.
Thus, their metallicity must be uniform, since these stars were born from the same gas cloud and at the same time stage.
However, to take this last property into account, we are faced with the drawback that this parameter is poorly reported in Gaia DR2 database.

@@

The main aim of this thesis is to \emph{build an unsupervised clustering model for open cluster characterization}.
The model will be \emph{non-supervised and non-parameterized} to fit a wide range of clusters without
the need for fine-tuning a high number of hyperparameters.

@@

\section{Comparing and Validating}
